Cultural Transmission Can Explain the Evolution of Cooperation

### 2020-08-10 ms/main.tex
conflict solving

%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Introduction
\section*{Introduction}
Cooperative behavior can harm an individual's fitness and increase the fitness of its conspecifics or competitors~\citep{axelrod1981evolution}.
Nevertheless, cooperative behavior appears to occur in many non-human animals~\citep{dugatkin1997cooperation}, for example rats~\citep{rice1962altruism} and birds~\citep{krams2008experimental}.
Evolution of cooperative behavior remains an important conundrum in evolutionary biology.

\emph{Kin selection} theory posits that natural selection can favor cooperation between related individuals. The importance of relatedness to the evolution of cooperation and altruism was shown by~\citet{hamilton1964genetical}. According to Hamilton, for an allele that determines cooperative behavior to increase in frequency, the reproductive cost to the actor that cooperates, $c$, must be less than the benefit to the recipient, $b$, times the `relatedness' between the recipient and the actor, $r$. This `relatedness' coefficient $r$ measures the correlation between the gene in the actor and the gene in the recipient.
This condition is also known as Hamilton's rule:
\begin{equation} \label{eq:hamilton_rule}
c < b \cdot r.
\end{equation}

\citet{Eshel1982} have studied a relevant model for the evolution of cooperative behavior under vertical transmission. Their model included \emph{assortative meeting}, or non-random encounters. That is, if a fraction $m$ of the population interacts with an individual of the same phenotype, and $1-m$ interacts randomly. 
Such assortative meeting may be due, for example, to population structure or active partner choice.
In their model, cooperative behavior can evolve if \footnote{In an extended model, which allows an individual to encounter $N$ individuals before choosing a partner, the righthand side is multiplied by $E[N]$, the expected number of encounters \citep[eq.~4.6]{Eshel1982}.
}. 
\citep[eq.~3.2]{Eshel1982}
\begin{equation} \label{eq:assortative_meeting}
c < b \cdot m,
\end{equation}
where $b$ and $c$ are the benefit and cost of cooperation
Here, $m$ takes the role of the relatedness $r$.

%These theories assume that cooperation is genetically determined, which raises the question: \emph{Is it possible that cooperation is determined by non-genetic factors?}
%Culture has significant impact on the behavior of humans~\citep{ihara2004cultural,jeong2018bronze} as well as non-human animals~\citep{bonner2018evolution}.
Here we attempt to determine to what extent the evolution of cooperative behavior can be explained by \emph{cultural transmission},
which allows an individual to acquire attitudes and behavioral traits from other individuals in its social group through imitation, learning, or other modes of communication \citep{cavalli1981cultural,richerson2008not}.
\citet{feldman1985gene} introduced the first model for the evolution of altruism by cultural transmission.
They showed that under vertical (parent-to-offspring) cultural transmission, Hamilton's rule does not govern the evolution of parent-to-offspring or sib-to-sib altruism.

\emph{Non-vertical transmission} may be either horizontal or oblique: horizontal transmission occurs between individuals from the same generation, while oblique transmission occurs from adults to unrelated offspring. 
Evolution under either of these transmission models  can be be more rapid than under pure vertical transmission~\citep{cavalli1981cultural,ram2018evolution}.
\citet{lewin2017microbes} have demonstrated that non-vertical transmission, mediated by microbes that manipulate their host behavior, can help to explain the evolution of cooperative behavior. Interestingly, some of their analysis can be applied to cultural transmission, because models of cultural transmission are mathematically similar to those for transmission of infectious diseases~\citep{cavalli1981cultural}.

We hypothesize that non-vertical cultural transmission can help explain the evolution of cooperation. 
To test this hypothesis, we suggest a model in which behavioral changes are mediated by cultural transmission that can occur during social interactions. For example, if an individual interacts with a cooperative individual, it might learn that cooperation is a positive behavior and will be cooperative in the future. 
We develop cultural evolution models that include both vertical and non-vertical transmission of cooperation and investigate these models using mathematical analysis and simulations.  
Our results demonstrate cultural transmission can facilitate the evolution of cooperation even when genetic transmission cannot.
These results suggest that further research on the evolution of cooperation should account for non-vertical transmission and that treatment of cooperation as a cultural, rather then genetic trait, can lead to a better understanding of this important and enigmatic phenomenon.

@@

We focus on the evolution of cooperation in a fully mixed population where cooperation is modeled using the \emph{prisoner's dilemma}. % yr: add ref

@@

Not all adults become parents due to natural selection, and we denote the frequency of phenotype $A$ among parents with $\tilde{p}$.
Therefore, the frequency $\hat{p}$ of  phenotype $A$ among juveniles (after selection and vertical and oblique transmission) is

@@

Individuals interact according to a prisoner's dilemma.
Specifically, individuals interact in pairs; a cooperator suffers a fitness cost $0<c<1$, and its partner gains a fitness benefit $b$, where we assume $c<b$. \textbf{\autoref{table:prisoner_payoff}} shows the payoff matrix, i.e. the fitness of an individual with phenotype $\phi_1$ when interacting with a partner of phenotype $\phi_2$.

%%% Table: payoff matrix
\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
           & $\phi_2=A$ & $\phi_2=B$ \\ \cmidrule(r){1-3}
$\phi_1=A$ & $1+b-c$ & $1-c$ \\
$\phi_1=B$ & $1+b$   & $1$
\\ \bottomrule
\end{tabular}
\caption{\textbf{Payoff matrix for prisoner's dilemma.}
The fitness of phenotype $\phi_1$ when interacting with phenotype $\phi_2$. $A$ is a cooperative phenotype, $B$ is a defector phenotype, $b$ is the benefit gained by an individual interacting with a cooperator, and $c$ is the cost of cooperation. $b>c>0$.
}
\label{table:prisoner_payoff}
\end{table}
%%%

Social interactions occur randomly:
two individuals with phenotype $A$ interact with probability $\hat{p}^2$, two individuals with phenotype $B$ interact with probability $(1-\hat{p})^2$, and two individuals with different phenotypes interact with probability $2\hat{p}(1-\hat{p})$.

@@

@@

\autoref{eq:nextgen_parents} can be simplified to
\begin{equation}\label{eq:nextgen_parents_simplified}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) \big(1-(1-\hat{p})(1-\alpha)T_B)\big) \\
& + \hat{p}(1-\hat{p}) (1-c) \big(\hat{p}(1-\alpha)T_B+1-T_B\big) \\
& + (1-\hat{p})\hat{p} (1+b) \big(\hat{p}(1-\alpha) + \alpha\big) T_A \\
& + (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
\end{aligned}
\end{equation}

%%% Table: interactions
\begin{table}[]
\begin{tabular}{@{}llllll@{}}
\toprule
\multirow{2}{*}{Phenotype $\phi_1$} &
  \multirow{2}{*}{Phenotype $\phi_2$} &
  \multirow{2}{*}{Frequency} &
  \multirow{2}{*}{Fitness of $\phi_1$} &
  \multicolumn{2}{l}{$P(\phi_1=A)$ via horizontal transmission:} \\ \cmidrule(l){5-6} 
    &     &                      &         & from partner, $\alpha$ & from population, $(1-\alpha)$ \\ \cmidrule(r){1-6}
$A$ & $A$ & $\hat{p}^2$          & $1+b-c$ & 1                      & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$A$ & $B$ & $\hat{p}(1-\hat{p})$ & $1-c$   & $1-T_B$                & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$B$ & $A$ & $\hat{p}(1-\hat{p})$ & $1+b$   & $T_A$                  & $\hat{p} T_A$                 \\
$B$ & $B$ & $(1-\hat{p})^2$      & $1$     & $0$                    & $\hat{p} T_A$                 \\ \bottomrule
\end{tabular}
\caption{\textbf{Interaction frequency, fitness, and transmission probabilities.}}
\label{table:interactions}
\end{table}
%%%

@@

With only oblique and horizontal transmission, i.e. $v = 0$, \autoref{eq:horizontal} becomes $\hat{p}=p$ and \autoref{eq:nextgen_adults_slimpify} becomes % TODO check this, I changed the ref from nextgen_parents_simplified to nextgen_adults_slimpify
\begin{equation}  \label{eq:nextgen_parents_oblique_only}
p' = p^2 (T_B-T_A) + p (1+T_A-T_B) ,
\end{equation}

@@

Therefore, in the absence of vertical transmission, selection plays no role in the evolution of cooperation. Hence, cooperation will evolve if the cooperator phenotype has a horizontal transmission bias (see~{\autoref{fig:results_c}).

@@

With only vertical and horizontal transmission, i.e. $v=1$, \autoref{eq:horizontal} becomes
$\hat{p} =  \tilde{p}$,
and \autoref{eq:nextgen_parents_simplified} for the frequency of the cooperative phenotype among parents in the next generation $\tilde{p}'$ can be written as
\begin{equation} \label{eq:nextgen_parents_vertical_only} 
\begin{aligned}
\bar{w} \tilde{p}' 
& = \tilde{p}^2 (1+b-c) [1 - (1-\tilde{p}) (1-\alpha) T_B] \\
& + \tilde{p}(1-\tilde{p}) (1-c) [\tilde{p} (1-\alpha) T_B + 1 - T_B] \\
& + \tilde{p}(1-\tilde{p}) (1+b) [\tilde{p} (1-\alpha) + \alpha] T_A \\
& + (1-\tilde{p})^2 \tilde{p} (1-\alpha) T_A .
\end{aligned}
\end{equation}

The fixation of either cooperation or defection, 
$\tilde{p}=0$ and $ \tilde{p}=1$, are equilibria of \autoref{eq:nextgen_parents_vertical_only}, that is, they solve $\tilde{p}'= \tilde{p}$.
We therefore assume for the remainder of the analysis that $0<\tilde{p}<1$.

If $\alpha=1$, then $\tilde{p}'= \tilde{p}$ is reduced to
\begin{equation}
\tilde{p}(1-\tilde{p})\big[(1+b)T_A + (1-c)(1-T_B)-1\big] = 0,
\end{equation}
and there are no additional equilibria.

Therefore, for cooperation to take over the population (for $\tilde{p}=1$ to be globally stable) we require $\tilde{p}'>\tilde{p}$, that is,
\begin{equation}
  \tilde{p}^2 (1+b-c) + \tilde{p}(1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > \bar{w}\tilde{p} .
\end{equation}
We divide by $\tilde{p}$, set $\bar{w} = 1 + \tilde{p}(b-c)$, and rearrange to get
\begin{equation}
  (1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > 1 -\tilde{p} .
\end{equation}
Dividing by $(1-\tilde{p})$ we find that $\tilde{p}'>\tilde{p}$ if 
\begin{equation} \label{eq:vert_hori_alpha1_condition_proof}
  (1-c) (1 - T_B) + (1+b)T_A
  > 1
\end{equation}
\\

If $\alpha<1$, we want to determine a condition for $\tilde{p}'>\tilde{p}$. 
We divide \autoref{eq:nextgen_parents_vertical_only} by $\tilde{p}$ and set $\bar{w} = 1 + \tilde{p}(b-c)$ to get
\begin{equation}
\begin{aligned} 
  1 + \tilde{p}(b-c) < 
  & \, \tilde{p}(1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
Rearranging, we get
\begin{equation} 
\begin{aligned} 
  1 - \tilde{p} < 
  & - \tilde{p}(1+b-c)(1-\tilde{p}) (1-\alpha) T_B \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
Diving by $(1-\tilde{p})$ and rearranging so that free terms are on the left and terms with $\tilde{p}$ are on the right, we have
\begin{equation} 
\begin{aligned} 
  &1 - (1-\alpha) T_A - (1+b) \alpha T_A - (1 - T_B)(1-c)  < \\
   &\tilde{p}[ - (1+b-c) (1-\alpha) T_B 
   + (1-c) (1-\alpha) T_B
   + (1+b) (1-\alpha) T_A 
   - (1-\alpha) T_A].
\end{aligned}
\end{equation}
Simplifying, we find that $\tilde{p}'>\tilde{p}$ if and only if
\begin{equation} \label{eq:vert_hori_global_condition}
c(1-T_B) - b \alpha T_A - (T_A-T_B) < \tilde{p} \cdot b (1-\alpha) (T_A-T_B).
\end{equation}

Following the same steps to solve $\tilde{p}'=\tilde{p}$, we find that there can be a third, polymorphic equilibrium 
\begin{equation} \label{eq:vert_hori_equilibrium}
  \tilde{p}^* = 
  \frac{c(1-T_B) - b \alpha T_A - (T_A-T_B)}{b (1-\alpha) (T_A-T_B)} .
\end{equation} 
Note that this is a legitimate equilibrium only if $0<\tilde{p}^*<1$.

Note that all parameters are positive.
So, applying \autoref{eq:vert_hori_global_condition}, for $\tilde{p}'>\tilde{p}$ we require that either 
\begin{align} 
\label{eq:vert_hori_global_condition1}
T_A > T_B \quad&\text{and}\quad \tilde{p}>\tilde{p}^*,  \quad \text{or} \\
\label{eq:vert_hori_global_condition2}
T_A <T_B \quad&\text{and}\quad \tilde{p}<\tilde{p}^* .
\end{align}

We therefore have the following result and corollaries.\\

@@

We define the initial frequency as $\tilde{p}_0$ and the cost boundaries
\begin{equation} \label{eq:cost_boundaries}
\begin{aligned}
\gamma_1 = \frac{b \alpha T_A + (T_A - T_B)}{1-T_B}, \quad
\gamma_2 = \frac{b \alpha T_B + (1+b) (T_A - T_B)}{1-T_B}.
\end{aligned}
\end{equation}
Applying eqs.~\ref{eq:vert_hori_equilibrium}, \ref{eq:vert_hori_global_condition1}, and \ref{eq:vert_hori_global_condition2} we can summarize the possible outcomes (\autoref{fig:results2}):
%\begin{enumerate} % p* version
%\item Fixation of cooperation,
%	if $T_A>T_B$ and $0<\tilde{p}^*$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0>\tilde{p}^*$, or
%	if $T_A<T_B$ and $1<\tilde{p}^*$.
%\item Global fixation of defection,
%	if $T_A>T_B$ and $\tilde{p}^*>1$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0<\tilde{p}^*$, or
%	if $T_A<T_B$ and $\tilde{p}^*<0$
%\item Co-existence of both phenotypes at $\tilde{p}^*$, 
%	if $T_A<T_B$ and $0<\tilde{p}^*<1$.
%\end{enumerate}
\begin{enumerate} % gamma version
\item Fixation of cooperation,
	if $T_A>T_B$ and $c<\gamma_1$; or
	if $T_A>T_B$ and $\gamma_1<c<\gamma_2$ and $\tilde{p}_0>\tilde{p}^*$; or
	if $T_A<T_B$ and $c<\gamma_2$.
\item Fixation of defection,
	if $T_A>T_B$ and $\gamma_2<c$; or
	if $T_A>T_B$ and $\gamma_1<c<\gamma_2$ and $\tilde{p}_0<\tilde{p}^*$; or
	if $T_A<T_B$ and $\gamma_1<c$.
\item Coexistence of both phenotypes at $\tilde{p}^*$, 
	if $T_A<T_B$ and $\gamma_2<c<\gamma_1$.
\end{enumerate}

It is interesting to note that cooperation and defection can stably coexist if there his horizontal bias for defection and the cost of cooperation is large but not too large. The recurrence for this case is illustrated in \autoref{fig:coexistence}.

Much of the literature on evolution of cooperation focuses on conditions for cooperation to invade a population of defectors.
The next corollary deals with such a condition, followed by a corollary that deals with symmetric horizontal transmission, i.e. $T_A=T_B$.
\\

@@

\begin{corollary}[Condition for cooperation to increase from rarity]
If the initial frequency of the cooperative phenotype if very close to zero, $\tilde{p}_0 \approx 0$, then its frequency will increase if 
\begin{equation} \label{eq:unequal_transmission_from_rarity}
\begin{aligned}
T_A>T_B \;\; \text{and} \;\; c < \gamma_1, \quad \text{or} \quad
T_A<T_B \;\; \text{and} \;\; \gamma_2<c < \gamma_1. 
\end{aligned}
\end{equation} 
\end{corollary}

In general, these conditions cannot be formulated in the form of Hamilton's rule ($c<b\cdot r$) due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, we get the following corollary that does have the form of Hamilton's rule.\\

\begin{corollary}[Symmetric horizontal transmission]
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \frac{\alpha T}{1-T}.
\end{equation}
\end{corollary}
To verify, set $T_A=T_B$ in \autoref{eq:vert_hori_global_condition}.

This can be interpreted as a version of Hamilton's rule (\autoref{eq:hamilton_rule}), where $\alpha T/(1-T)$ is the `effective relatedness'.
\autoref{fig:results_a} demonstrates this condition. 
\\

\begin{corollary}[No assortment of transmission and cooperation]
When $\alpha=0$, then if there is horizontal bias for cooperation ($T_A>T_B$) and
\textbf{(1)} the cost is low compared to the bias ($c<(T_A-T_B)/(1-T_B)$), then cooperation will fix from any positive frequency; or
\textbf{(2)} the cost is low compared to the benefit ($c<(1+b)(T_A-T_B)(1-T_B)$), then cooperation will fix if the initial frequency if high enough ($\tilde{p}_0 > \tilde{p}^*$).
\end{corollary}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Introduction
\section*{Introduction}
Cooperative behavior can reduce an individual's fitness and increase the fitness of its conspecifics or competitors~\citep{axelrod1981evolution}.
Nevertheless, cooperative behavior appears to occur in many non-human animals~\citep{dugatkin1997cooperation}, including primates~\citep{jaeggi2013natural},  rats~\citep{rice1962altruism}, birds~\citep{stacey1990cooperative,krams2008experimental}, and lizards~\citep{sinervo2006self}.
Evolution of cooperative behavior remains an important conundrum in evolutionary biology. Since the work of  \citet{hamilton1964genetical} and \citet{axelrod1981evolution}, theories for the evolution of cooperative and altruistic behaviors have been intertwined often under the rubric of kin selection.

\emph{Kin selection} theory posits that natural selection is more likely to favor cooperation between more closely related individuals. The importance of relatedness to the evolution of cooperation and altruism was shown by~\citet{hamilton1964genetical}, who showed that an allele that determines cooperative behavior will increase in frequency if the reproductive cost to the actor that cooperates, $c$, is less than the benefit to the recipient, $b$, times the relatedness, $r$, between the recipient and the actor. This relatedness coefficient $r$ measures a probability that an allele sampled from the cooperator is identical by descent to one at the same locus in the recipient.
This condition is  known as ``Hamilton's rule'':
\begin{equation} \label{eq:hamilton_rule}
c < b \cdot r.
\end{equation}

\citet{Eshel1982}  studied a related model for the evolution of cooperative behavior under vertical transmission. Their model included \emph{assortative meeting}, or non-random encounters, where a fraction $m$ of individuals in the population each interact with an individual of the same phenotype, and a fraction $1-m$ interacts  with a randomly chosen individual.  
Such assortative meeting may be due, for example, to population structure or active partner choice.
In their model, cooperative behavior can evolve if\footnote{In an extended model, which allows an individual to encounter $N$ individuals before choosing a partner, the righthand side is multiplied by $E[N]$, the expected number of encounters \citep[eq.~4.6]{Eshel1982}.
} 
\citep[eq.~3.2]{Eshel1982}
\begin{equation} \label{eq:assortative_meeting}
c < b \cdot m\,,
\end{equation}
where $b$ and $c$ are the benefit and cost of cooperation. 
Here $m$ in (\ref{eq:assortative_meeting}) takes the role of the relatedness $r$  in (\ref{eq:hamilton_rule}).

%These theories assume that cooperation is genetically determined, which raises the question: \emph{Is it possible that cooperation is determined by non-genetic factors?}
%Culture has significant impact on the behavior of humans~\citep{ihara2004cultural,jeong2018bronze} as well as non-human animals~\citep{bonner2018evolution}.

Here we study the evolution of a cooperative behavior that is subject to \emph{cultural transmission},
 which allows an individual to acquire attitudes or behavioral traits from other individuals in its social group through imitation, learning, or other modes of communication \citep{cavalli1981cultural,richerson2008not}.
\citet{feldman1985gene} introduced the first model for the evolution of altruism by cultural transmission and
showed that under vertical (parent-to-offspring) cultural transmission, the conditions for the initial increase of altruism entails a modification of Hamilton's rule in the cases  of parent-to-offspring or sib-to-sib altruism.

\emph{Non-vertical cultural transmission} may be either viewed as horizontal or oblique: horizontal transmission occurs between individuals from the same generation, while oblique transmission occurs  to  offspring from the generation to which their parents belong. 
Evolution under either of these transmission models  can be be more rapid than under pure vertical transmission~\citep{cavalli1981cultural,lycett2008questions,ram2018evolution}.
\citet{lewin2017microbes} have demonstrated that non-vertical transmission, mediated by microbes that manipulate their host's behavior, can help to explain the evolution of cooperative behavior. Some of their analysis can be applied to cultural transmission, because models of cultural transmission are mathematically similar to those for transmission of infectious diseases~\citep{cavalli1981cultural}.

We hypothesize that non-vertical cultural transmission can enhance the evolution of cooperation, and 
to test this hypothesis we suggest a model in which behavioral changes are mediated by cultural transmission that can occur during social interactions. For example, if an individual interacts with a cooperative individual, it might learn that cooperation is an advantageous behavior and will be cooperative in the future. 
Our cultural evolution models  include both vertical and non-vertical transmission of cooperation, and we investigate these models using mathematical analysis and simulations.  
Our results demonstrate that cultural transmission can facilitate the evolution of cooperation even when genetic transmission can not, and that treatment of cooperation as a cultural, rather than a genetic, trait can lead to a better understanding of its evolutionary dynamics.

@@

@@

Not all adults become parents due to natural selection, and we denote the frequency of phenotype $A$ among parents by $\tilde{p}$.
Therefore, the frequency $\hat{p}$ of  phenotype $A$ among juveniles (after selection and vertical and oblique transmission) is

@@

Individuals are assumed to interact according to a \emph{prisoner's dilemma}.
Specifically, individuals interact in pairs; a cooperator suffers a fitness cost $0<c<1$, and its partner gains a fitness benefit $b$, where we assume $c<b$. \textbf{\autoref{table:prisoner_payoff}} shows the payoff matrix, i.e. the fitness of an individual with phenotype $\phi_1$ when interacting with a partner of phenotype $\phi_2$.

\bigskip
\begin{center}[TABLE 1 HERE]\end{center}
%%% Table: payoff matrix

\bigskip
Social interactions occur randomly:
two individuals with phenotype $A$ interact with probability $\hat{p}^2$, two individuals with phenotype $B$ interact with probability $(1-\hat{p})^2$, and two individuals with different phenotypes interact with probability $2\hat{p}(1-\hat{p})$.

@@

\bigskip
\begin{center}[TABLE 2 HERE]\end{center}
%%% Table: payoff matrix

\bigskip

@@

\autoref{eq:nextgen_parents} can be simplified to
\begin{equation}\label{eq:nextgen_parents_simplified}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) \big(1-(1-\hat{p})(1-\alpha)T_B)\big) \\
& + \hat{p}(1-\hat{p}) (1-c) \big(\hat{p}(1-\alpha)T_B+1-T_B\big) \\
& + (1-\hat{p})\hat{p} (1+b) \big(\hat{p}(1-\alpha) + \alpha\big) T_A \\
& + (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
\end{aligned}
\end{equation}
where $\hat{p}=v\tilde{p}+(1-v)p$.
%This is not a recursion-MF

%%% Table: interactions

@@

With only oblique and horizontal transmission, i.e. $v = 0$, Eq.\ \ref{eq:horizontal} becomes $\hat{p}=p$ and Eq.\ \ref{eq:nextgen_adults_slimpify} becomes % TODO check this, I changed the ref from nextgen_parents_simplified to nextgen_adults_slimpify
\begin{equation}  \label{eq:nextgen_parents_oblique_only}
p' = p^2 (T_B-T_A) + p (1+T_A-T_B) ,
\end{equation}

@@

Therefore, in the absence of vertical transmission, selection plays no role in the evolution of cooperation. Hence, cooperation will evolve if the cooperator phenotype has a horizontal transmission bias (see~{\autoref{fig:results}c).

@@

With only vertical and horizontal transmission, i.e. $v=1$, Eq.\ \ref{eq:horizontal} becomes
$\hat{p} =  \tilde{p}$,
and Eq.\ \ref{eq:nextgen_parents_simplified} for the frequency of the cooperative phenotype among parents in the next generation $\tilde{p}'$ can be written as
\begin{equation} \label{eq:nextgen_parents_vertical_only} 
\begin{aligned}
\bar{w} \tilde{p}' 
& = \tilde{p}^2 (1+b-c) [1 - (1-\tilde{p}) (1-\alpha) T_B] \\
& + \tilde{p}(1-\tilde{p}) (1-c) [\tilde{p} (1-\alpha) T_B + 1 - T_B] \\
& + \tilde{p}(1-\tilde{p}) (1+b) [\tilde{p} (1-\alpha) + \alpha] T_A \\
& + (1-\tilde{p})^2 \tilde{p} (1-\alpha) T_A .
\end{aligned}
\end{equation}

Fixation of either cooperation, $\tilde{p}=1$, or defection, 
$\tilde{p}=0$, are equilibria of Eq.\ \ref{eq:nextgen_parents_vertical_only}, that is, they solve $\tilde{p}'= \tilde{p}$.
We therefore assume for the remainder of the analysis that $0<\tilde{p}<1$.

If $\alpha=1$, then $\tilde{p}'= \tilde{p}$ is reduced to
\begin{equation}
\tilde{p}(1-\tilde{p})\big[(1+b)T_A + (1-c)(1-T_B)-1\big] = 0,
\end{equation}
and there are no additional equilibria.
For cooperation to take over the population (for $\tilde{p}=1$ to be globally stable) we require $\tilde{p}'>\tilde{p}$; that is,
\begin{equation}
  \tilde{p}^2 (1+b-c) + \tilde{p}(1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > \bar{w}\tilde{p} .
\end{equation}
We divide by $\tilde{p}$, set $\bar{w} = 1 + \tilde{p}(b-c)$, and rearrange to get
\begin{equation}
  (1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > 1 -\tilde{p} .
\end{equation}
Dividing by $(1-\tilde{p})$ we find that $\tilde{p}'>\tilde{p}$ if 
\begin{equation} \label{eq:vert_hori_alpha1_condition_proof}
  (1-c) (1 - T_B) + (1+b)T_A
  > 1
\end{equation}
\\

If $\alpha<1$, divide both sides of Eq.\ \ref{eq:nextgen_parents_vertical_only} by $\tilde{p}$ and set $\bar{w} = 1 + \tilde{p}(b-c)$. Then $\tilde{p}'>\tilde{p}$ if
\begin{equation}
\begin{aligned} 
  1 + \tilde{p}(b-c) < 
  & \, \tilde{p}(1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
Rearranging, we get
\begin{equation} 
\begin{aligned} 
  1 - \tilde{p} < 
  & - \tilde{p}(1+b-c)(1-\tilde{p}) (1-\alpha) T_B \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
Diving by $(1-\tilde{p})$ and rearranging so that free terms are on the left and terms with $\tilde{p}$ are on the right, we have
\begin{equation} 
\begin{aligned} 
  &1 - (1-\alpha) T_A - (1+b) \alpha T_A - (1 - T_B)(1-c)  < \\
   &\tilde{p}[ - (1+b-c) (1-\alpha) T_B 
   + (1-c) (1-\alpha) T_B
   + (1+b) (1-\alpha) T_A 
   - (1-\alpha) T_A].
\end{aligned}
\end{equation}
Simplifying, we find that $\tilde{p}'>\tilde{p}$ if and only if
\begin{equation} \label{eq:vert_hori_global_condition}
c(1-T_B) - b \alpha T_A - (T_A-T_B) < \tilde{p} \cdot b (1-\alpha) (T_A-T_B).
\end{equation}

Besides the fixation states $\tilde{p}=0$ and $\tilde{p}=1$, there may be an actual polymorphic equilibrium of $\tilde{p}'=\tilde{p}$ in Eq.\ \ref{eq:nextgen_parents_vertical_only}, namely 
\begin{equation} \label{eq:vert_hori_equilibrium}
  \tilde{p}^* = 
  \frac{c(1-T_B) - b \alpha T_A - (T_A-T_B)}{b (1-\alpha) (T_A-T_B)} ,
\end{equation} 
which is legitimate if $0<\tilde{p}^*<1$.

Since all parameters are positive, we can apply
inequality \ref{eq:vert_hori_global_condition} and see that for $\tilde{p}'>\tilde{p}$ we require that either 
\begin{align} 
\label{eq:vert_hori_global_condition1}
T_A > T_B \quad&\text{and}\quad \tilde{p}>\tilde{p}^*,  \quad \text{or} \\
\label{eq:vert_hori_global_condition2}
T_A <T_B \quad&\text{and}\quad \tilde{p}<\tilde{p}^* .
\end{align}

We summarize these findings in the following result and corollaries.\\

@@

Let the initial frequency of the alternative phenotype be $\tilde{p}_0$ and denote the cost boundaries by
\begin{equation} \label{eq:cost_boundaries}
\begin{aligned}
\gamma_1 = \frac{b \alpha T_A + (T_A - T_B)}{1-T_B}, \quad
\gamma_2 = \frac{b \alpha T_B + (1+b) (T_A - T_B)}{1-T_B}.
\end{aligned}
\end{equation}
Then, applying \ref{eq:vert_hori_equilibrium}, \ref{eq:vert_hori_global_condition1}, and \ref{eq:vert_hori_global_condition2}, we  summarize the possible outcomes which are illustrated in \autoref{fig:results2}:
%\begin{enumerate} % p* version
%\item Fixation of cooperation,
%	if $T_A>T_B$ and $0<\tilde{p}^*$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0>\tilde{p}^*$, or
%	if $T_A<T_B$ and $1<\tilde{p}^*$.
%\item Global fixation of defection,
%	if $T_A>T_B$ and $\tilde{p}^*>1$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0<\tilde{p}^*$, or
%	if $T_A<T_B$ and $\tilde{p}^*<0$
%\item Co-existence of both phenotypes at $\tilde{p}^*$, 
%	if $T_A<T_B$ and $0<\tilde{p}^*<1$.
%\end{enumerate}
\begin{enumerate} % gamma version
\item Fixation of cooperation,
	if $T_A>T_B$ and $c<\gamma_1$; or
	if $T_A>T_B$ and $\gamma_1<c<\gamma_2$ and $\tilde{p}_0>\tilde{p}^*$; or
	if $T_A<T_B$ and $c<\gamma_2$.
\item Fixation of defection,
	if $T_A>T_B$ and $\gamma_2<c$; or
	if $T_A>T_B$ and $\gamma_1<c<\gamma_2$ and $\tilde{p}_0<\tilde{p}^*$; or
	if $T_A<T_B$ and $\gamma_1<c$.
\item Coexistence of both phenotypes at $\tilde{p}^*$, 
	if $T_A<T_B$ and $\gamma_2<c<\gamma_1$.
\end{enumerate}

Note that cooperation and defection can  coexist stably if there is horizontal bias for defection and the cost of cooperation is large but not too large. The recursion dynamic for this case is illustrated in \autoref{fig:coexistence}.

Much of the literature on evolution of cooperation focuses on conditions for an initially rare cooperative phenotype to invade a population of defectors.
The next corollary deals with such a condition, followed by a corollary that deals with symmetric horizontal transmission, i.e. $T_A=T_B$.
\\

@@

\begin{corollary}[Condition for cooperation to increase when initially rare]
If the initial frequency of the cooperative phenotype is very close to zero, $\tilde{p}_0 \approx 0$, then this frequency will increase if 
\begin{equation} \label{eq:unequal_transmission_from_rarity}
\begin{aligned}
T_A>T_B \;\; \text{and} \;\; c < \gamma_1, \quad \text{or} \quad
T_A<T_B \;\; \text{and} \;\; \gamma_2<c < \gamma_1. 
\end{aligned}
\end{equation} 
\end{corollary}

In general, these conditions cannot be formulated in the form of Hamilton's rule ($c<b\cdot r$) due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T_A=T_B$, these conditions reduce to the following  form of Hamilton's rule.\\

\begin{corollary}[Symmetric horizontal transmission]
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \frac{\alpha T}{1-T}.
\end{equation}
\end{corollary}
Inequality \ref{eq:equal_transmission} is obtained by setting $T_A=T_B$ in inequality \ref{eq:vert_hori_global_condition} and can be interpreted as a version of Hamilton's rule (inequality \ref{eq:hamilton_rule}), where $\alpha T/(1-T)$ can be regarded as the `effective relatedness'.
\autoref{fig:results}a demonstrates this condition. 
\\
